Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-29 paper.tex

\textbf{Results:}
Zarr is a format for storing
multi-dimensional data that is widely used across the sciences,
and is ideally suited to massively parallel processing.
We present the VCF Zarr specification, an encoding of the
VCF data model using Zarr, along with fundamental software infrastructure
for efficient and reliable conversion at scale.
We show how this format is far more efficient than
standard VCF based approaches,
and competitive with specialised methods for
storing genotype data in terms of compression ratios
and single-threaded calculation performance.
We present case studies on subsets of three large 
human datasets (Genomics England: $n$=78,195;
Our Future Health: $n$=651,050; 
All of Us: $n$=245,394) along with whole genome
datasets for Norway Spruce ($n$=1,063) and 
SARS-Cov-2 ($n$=3,960,704), % TODO update for tranche 2 data 
demonstrating the scalability and flexibility of VCF Zarr.
We demonstrate the potential for VCF Zarr to enable a new generation
of high-performance and cost-effective applications
via illustrative highly parallel benchmarks on AWS and experimentation with GPUs.

@@

In this article, we decouple the VCF data model from its row-oriented
file definition, and show how the data can be
compactly stored and efficiently analysed in a cloud-native, FAIR manner.
We do this by translating VCF data into Zarr format,
a method of storing large-scale multidimensional data as a regular
grid of compressed chunks.
Zarr's elegant simplicity and first-class support for
cloud object stores and parallel computing have led to
it gaining substantial traction
across the sciences, and it is now used in multiple petabyte-scale
datasets in cloud deployments (see Methods for details).
We present the VCF Zarr specification that formalises this
mapping, along with the \texttt{vcf2zarr} and \texttt{vcztools} 
utilities to reliably convert VCF to and from Zarr at scale.
We show that VCF Zarr is much more compact than
VCF and is competitive with state-of-the-art
file-based VCF compression tools.
Moreover, we show that Zarr's storage of data in an analysis-ready
format greatly facilitates computation,
with various benchmarks being substantially faster than
\texttt{bcftools} based pipelines, and again competitive
with state-of-the-art file-oriented methods. 
We demonstrate the scalability and flexibility of VCF Zarr
via case-studies on five boundary-pushing datasets.
We examine conversion and compression performance on three different
data modalities for large human datasets: whole genome sequence
data from Genomics England (5X smaller than gzipped VCF);
genotype data from Our Future Health (2.5X smaller than gzipped VCF);
and exome-like data from All of Us (1.1X larger than gzipped VCF). 
We then examine performance 
for species with large genomes using Norway Spruce data (3.2X smaller
than gzipped VCF),
and in a large collection of short whole genome alignments
using SARS-Cov-2 data 
(100X % FIXME! this is roughly correct, but need to get the numbers straight
smaller than gzipped FASTA). 
Interoperability with the vast ecosystem of tools
based on VCF~\citep{danecek2021twelve,garrison2022spectrum}
is crucial. 
% TODO sentence too long
We first present the \texttt{vcztools} program,
which implements a subset of the functionality in \texttt{bcftools}
and allows data stored in Zarr at some URL to be 
treated as if it were a VCF or BCF file,
and demonstrate that it can be used as a practical replacement 
in processing pipelines.
We then present an illustrative example of how existing
methods can adopt VCF Zarr by modifying the popular 
SAIGE~\cite{zhou2018efficiently, zhou2020scalable} GWAS software
(written in C++), and show significant speedups over the existing 
VCF (6.5X) and BCF (3X) backends in a simple benchmark.
Finally, we illustrate the potential 
for a new generation of Zarr-based applications via some illustative
benchmarks. First, we demonstrate the very high levels of throughput 
performance available on public clouds when data in an object 
store is accessed in an asynchronous and highly parallel manner.
Finally, we demonstrate the ease and efficiency with which data
in Zarr format can be processed on a GPU, opening many exiting
avenues for future research.

@@

\subsection{Case study: SARS-Cov-2 alignment data}
The SARS-CoV-2 pandemic resulted in viral sequence data collection at
an unprecedented scale, with over 15 million whole genomes incorporated into
one resource as of 2024~\citep{hinrichs2024ongoing}. Data at such scale
overwhelmed classical phylogenetic methods, leading to the development
of a new class of tools~\citep{turakhia2021ultrafast,ye2022matoptimize,zhan2023towards}.
The Viridian dataset~\citep{hunt2024addressing} is a consistently assembled
set of consensus sequences for 4,471,579 samples, addressing a range of systematic
errors in earlier collections. The \texttt{sc2ts} inference tool uses VCF-Zarr as its
input format, and Zhan et al have made the Viridian dataset,
aligned to the reference using MAFFT~\citep{katoh2013mafft},
available for download~\citep{zhan2023towards}. This dataset has quite
different properties to those examined in previous case studies, and
illustrates the flexibility of Zarr across a range of biological applications.

>>>>

\textbf{Results:}
Zarr is a format for storing
multi-dimensional data that is widely used across the sciences,
and is ideally suited to massively parallel processing.
We present the VCF Zarr specification, an encoding of the
VCF data model using Zarr, along with fundamental software infrastructure
for efficient and reliable conversion at scale.
We show how this format is far more efficient than
standard VCF based approaches,
and competitive with specialised methods for
storing genotype data in terms of compression ratios
and single-threaded calculation performance.
We present case studies on subsets of three large 
human datasets (Genomics England: $n$=78,195;
Our Future Health: $n$=651,050; 
All of Us: $n$=245,394) along with whole genome
datasets for Norway Spruce ($n$=1,063) and 
SARS-CoV-2 ($n$=3,960,704), % TODO update for tranche 2 data 
demonstrating the scalability and flexibility of VCF Zarr.
We demonstrate the potential for VCF Zarr to enable a new generation
of high-performance and cost-effective applications
via illustrative highly parallel benchmarks on AWS and experimentation with GPUs.

@@

In this article, we decouple the VCF data model from its row-oriented
file definition, and show how the data can be
compactly stored and efficiently analysed in a cloud-native, FAIR manner.
We do this by translating VCF data into Zarr format,
a method of storing large-scale multidimensional data as a regular
grid of compressed chunks.
Zarr's elegant simplicity and first-class support for
cloud object stores and parallel computing have led to
it gaining substantial traction
across the sciences, and it is now used in multiple petabyte-scale
datasets in cloud deployments (see Methods for details).
We present the VCF Zarr specification that formalises this
mapping, along with the \texttt{vcf2zarr} and \texttt{vcztools} 
utilities to reliably convert VCF to and from Zarr at scale.
We show that VCF Zarr is much more compact than
VCF and is competitive with state-of-the-art
file-based VCF compression tools.
Moreover, we show that Zarr's storage of data in an analysis-ready
format greatly facilitates computation,
with various benchmarks being substantially faster than
\texttt{bcftools} based pipelines, and again competitive
with state-of-the-art file-oriented methods. 
We demonstrate the scalability and flexibility of VCF Zarr
via case-studies on five boundary-pushing datasets.
We examine conversion and compression performance on three different
data modalities for large human datasets: whole genome sequence
data from Genomics England (5X smaller than gzipped VCF);
genotype data from Our Future Health (2.5X smaller than gzipped VCF);
and exome-like data from All of Us (1.1X larger than gzipped VCF). 
We then examine performance 
for species with large genomes using Norway Spruce data (3.2X smaller
than gzipped VCF),
and in a large collection of short whole genome alignments
using SARS-CoV-2 data 
(100X % FIXME! this is roughly correct, but need to get the numbers straight
smaller than gzipped FASTA). 
Interoperability with the vast ecosystem of tools
based on VCF~\citep{danecek2021twelve,garrison2022spectrum}
is crucial. 
% TODO sentence too long
We first present the \texttt{vcztools} program,
which implements a subset of the functionality in \texttt{bcftools}
and allows data stored in Zarr at some URL to be 
treated as if it were a VCF or BCF file,
and demonstrate that it can be used as a practical replacement 
in processing pipelines.
We then present an illustrative example of how existing
methods can adopt VCF Zarr by modifying the popular 
SAIGE~\cite{zhou2018efficiently, zhou2020scalable} GWAS software
(written in C++), and show significant speedups over the existing 
VCF (6.5X) and BCF (3X) backends in a simple benchmark.
Finally, we illustrate the potential 
for a new generation of Zarr-based applications via some illustative
benchmarks. First, we demonstrate the very high levels of throughput 
performance available on public clouds when data in an object 
store is accessed in an asynchronous and highly parallel manner.
Finally, we demonstrate the ease and efficiency with which data
in Zarr format can be processed on a GPU, opening many exiting
avenues for future research.

@@

\subsection{Case study: SARS-CoV-2 alignment data}
The SARS-CoV-2 pandemic resulted in viral sequence data collection at
an unprecedented scale, with over 15 million whole genomes incorporated into
one resource as of 2024~\citep{hinrichs2024ongoing}. Data at such scale
overwhelmed classical phylogenetic methods, leading to the development
of a new class of tools~\citep{turakhia2021ultrafast,ye2022matoptimize,zhan2023towards}.
The Viridian dataset~\citep{hunt2024addressing} is a consistently assembled
set of consensus sequences for 4,471,579 samples, addressing a range of systematic
errors in earlier collections. The \texttt{sc2ts} inference tool uses VCF-Zarr as its
input format, and Zhan et al have made the Viridian dataset,
aligned to the reference using MAFFT~\citep{katoh2013mafft},
available for download~\citep{zhan2023towards}. This dataset has quite
different properties to those examined in previous case studies, and
illustrates the flexibility of Zarr across a range of biological applications.
